Simple repression draft

### 2020-06-05 doc/section_04_bayesian_inference.tex
Not quite final pass on everything, Appdx S3 still needs more

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
constitutive promoters, achieves the right balance of complexity and simplicity,
by allowing both Fano factor $\nu>1$, but also by remedying, by design, the
problems of parameter degeneracy than model 4 in
Figure~\ref{fig2:constit_cartoons} suffered~\cite{Razo-Mejia2020}. Does this
stand up to closer scrutiny, namely, comparison to full mRNA distributions
rather than simply their moments? We will test this thoroughly on single-cell
mRNA counts for different unregulated promoters from Jones et.\
al.~\cite{Jones2014}.

@@

For this model the master equation of interest is
Eq.~\ref{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
\begin{equation}
{d\over dt}p_{m,U}(t) = 
        rp_{m-1,U}(t) 
        - rp_{m,U}(t)
        + (m+1)\gamma p_{m+1,U}(t) 
        - \gamma p_{m,U}(t),
\end{equation}
whose steady-state solution is given by a Poisson distribution with parameter
$\lambda \equiv r / \gamma$~\cite{Sanchez2013}. The goal of our inference 
problem is then to find the probability distribution for the parameter value
$\lambda$ given the experimental data. By Bayes theorem this can be written as
\begin{equation}
p(\lambda \mid D) = {p(D \mid \lambda) p(\lambda) \over p(D)},
\end{equation}
where $D = \{m_1, m_2, \ldots, m_N \}$ are the single-cell mRNA experimental
counts. As is standard we will neglected the denominator $p(D)$ on the right
hand side since it is independent of $\lambda$ and serves only as a
normalization factor.

The steady-state solution for the master equation defines the likelihood term
for a single cell $p(m \mid \lambda)$. What this means is that for a given
choice of parameter $\lambda$, under model 1 of
Figure~\ref{fig2:constit_cartoons}(A), we expect to observe $m$ mRNAs in a
single cell with probability
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}.
\label{eq:poisson_inference010}
\end{equation}
Assuming each cell's mRNA count in our dataset is independent of others, the
likelihood of the full inference problem $p(D\mid\lambda)$ is simply a product
of the single cell likelihoods given by \eq{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}

@@

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is
known as the \textit{conjugate} prior for a Poisson likelihood. This type of
priors have nice properties to work with, in particular a conjugate prior allows
us to get a closed form for the posterior distribution $p(\lambda \mid D)$. For
our case it can be shown that for a Poisson distribution likelihood with its
Gamma distribution conjugate prior, the posterior distribution is also a Gamma
distribution ~\cite{Gelman2013}. Specifically the two parameters $\alpha'$ and
$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
\frac{1}{N}\sum_{k=1}^N m_k$ for notational convenience, and $N$ is the number
of cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to approximate the
Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
we have inferred our model's one parameter to a precision of order 1\%.

We remind the reader that we began this section claiming that the Poisson model
was ``wrong'' since it could not reproduce features of the data such as a Fano
factor $> 1$. The fact that we got such a narrow posterior distribution for our
parameter $P(\lambda \mid D)$ does not equate to the model being adequate to
describe the data. What this means is that given the data $D$, the parameter 
$\lambda$ for our model is highly constrained to take a narrow range of values,
but a narrow posterior distribution does not speak to the goodness of fit of the
model. As we will see later in Figure~\ref{fig:constit_post_full} after
exploring the bursty promoter model, indeed there is not good correspondence
when contrasting the Poisson model with the experimental data.

@@

Let us now consider the problem of parameter inference for model five
from~\fig{fig1:means_cartoons}(C). As derived in
Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
model is a negative binomial distribution, given by
\begin{equation}
p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
        \left(\frac{1}{1+b}\right)^{k_i}
        \left(\frac{b}{1+b}\right)^m,
\label{eq:neg_bionom}
\end{equation}
where $b$ is the mean burst size and $k_i$ is the burst rate in units of the
mRNA degradation rate $\gamma$. As sketched earlier, to think of the negative
binomial distribution in terms of an intuitive story, we imagine the arrival of
bursts as a Poisson process with rate $k_i$, where each burst has a
geometrically-distributed size with mean size $b$.

@@

We carried out Markov-Chain Monte Carlo (MCMC) sampling on the posterior of this
model, starting with the constitutive \textit{lacUV5} dataset
from~\cite{Jones2014}. The resulting MCMC samples are shown in
Figure~\ref{fig:constit_post_full}(A). In contrast to the active/inactive
constitutive model considered in~\cite{Razo-Mejia2020} (nonequilibrium model 4
in~\ref{fig2:constit_cartoons}), this model is well-identified with both
parameters determined to a fractional uncertainty of 5-10\%. The strong
correlation reflects the fact that their product sets the mean of the mRNA
distribution, which is tightly constrained by the data, but there is weak
``sloppiness''~\cite{Transtrum2015} along a set of values with a similar
product.

@@

\fig{fig:constit_post_full}(C) shows the results, plotting the posterior
distribution from individually MCMC sampling all 18 constitutive promoter
datasets from~\cite{Jones2014}. To aid visualization, rather than plotting
samples for each promoter's posterior as in \fig{fig:constit_post_full}(A), for
each posterior we find and plot the curve that surrounds the 95\% highest
probability density region. What this means is that each contour 
encloses approximately 95\% of the samples, and thus 95\% of the probability
mass, of its posterior distribution. Theory-experiment comparisons (not shown)
display a similar level of agreement between data and predictive samples as for
the bursty model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).
\mmnote{Might be worth showing some/all of these PPCs in supplement for
completeness? We could do postage stamp sized diff plots and probably fit all 18
promoters on 1-2 pages.}\mrm{Definitely need to show them in SI.}

@@

\mmnote{Outline, still to cover:
\begin{itemize}
\item There is a puzzle in comparing w/ Chong2014 supercoiling
model: if supercoiling is the thing, why are my burst sizes all
the same but burst rates vary? And why is the duty cycle of my
promoter so low, ie., bursts so short? COmpare their fig 7E;
their $\beta/\alpha$ is my $k^+/k^-$. They have one or two genes
with very small $\beta/\alpha$, does the \textit{galK} locus just
happen to be that, or is there a deeper disagreement? Hard to say
w/o more data. Are we
missing important things?? Unclear, we leave as open question. Good enough for
our purposes: posterior is identifiable, PPC is great, and of the models we've
thought of it is unique in satisfying both.
\end{itemize}
}

@@

Before showing the mathematical formulation of our statistical inference model,
we would like to sketch the intuitive structure. The dataset
from~\cite{Jones2014} we consider consists of single-cell mRNA counts data of
nine different conditions, spanning several combinations of three unique
repressor binding sites and four unique repressor copy numbers. We assume that
the values of $k_i$ and $b$ are known, since we have already cleanly inferred
them from constitutive promoter data, and further we assume that these values
are the same across datasets with different repressor binding sites and copy
numbers. In other words, we assume that the regulation of the transcription
factor does not affect the mean burst size nor the burst initiation rate. The
regulation happens as the promoter is taken away from the transcriptionally
active state when the promoter is bound by repressor. We assume that there is
one unbinding rate parameter for each repressor binding site, and likewise one
binding rate for each unique repressor copy number. This makes our model seven
dimensional, or nine if one counts $k_i$ and $b$ as well. Note that we use only
a subset of the datasets from Jones et.\ al.~\cite{Jones2014}, as discussed more
in Appendix~\ref{sec:bayes}.

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\},
\end{equation}
where subscripts for dissociation rates $k^-$  indicate the different repressor
binding sites, and subscripts for association rates $k^+$ indicate the
concentration of the small-molecule that controlled the expression of the LacI
repressor (see Appendix~\ref{sec:bayes}). This is because for this particular
dataset the repressor copy numbers were not measured directly, but it is safe to
assume that a given concentration of the inducer resulted in a specific mean
repressor copy number\cite{Chure2019a}. Also note that the authors
of~\cite{Jones2014} report estimates of LacI copy number per cell rather than
direct measurements. However, these estimates were made assuming the validity of
the equilibrium models in Figure~\ref{fig1:means_cartoons}, and since testing
these models is our present goal, it would be circular logic if we were to make
the same assumption. Therefore we will make no assumptions about the LacI copy
number for a given inducer concentrations.

Having stated the problem, Bayes theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is again the set of all $N$ observed single-cell mRNA counts
across the various conditions. We assume that individual single-cell
measurements are independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding
rates out of $\vect{k}$ for the $j$-th measured cell. The probability
$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
distribution for our bursty model with repression derived in
Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $_2F_1$ is the confluent hypergeometric function of the second kind,
$\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

This likelihood is rather inscrutable. We did not find any of the known
analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
we instead resorted to numerics. One insight we found
\mmnote{a plot would probably help explain, maybe SI, need to explain better}
was that for very strong or very weak repression, the distribution in
Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
values, except with $k_i$ multiplied by the fold-change
$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the ratio
$k_R^+/k_R^-$ was detectable. But for intermediate repression, the distribution
was visibly broadened with Fano factor greater than $1+b$, the value for the
corresponding constitutive case. This indicates that the repressor rates had
left an imprint on the distribution, and perhaps intuitively, this intermediate
regime occurs for values of $k_R^\pm$ comparable to the burst rate $k_i$. Put
another way, if the repressor rates are much faster or much slower than $k_i$,
then there is a timescale separation and effectively only one timescale remains,
$k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three rates in the problem
are comparable does the mRNA distribution retain detectable information about
them.

Next we specify priors. As for the constitutive model, weakly informative
lognormal priors are a natural choice for all our rates. We found that if the
priors were too weak, our MCMC sampler would often become stuck in regions of
parameter space with very low probability density, unable to move. We struck a
balance in choosing our prior widths between helping the sampler run while
simultaneously verifying that the marginal posteriors for each parameter were
not artificially constrained or distorted by the presence of the prior. The only
exception to this is the highly informative priors we placed on $k_i$ and $b$,
since we have strong knowledge of them from our inference of constitutive
promoters above. All details for our prior distributions are listed in 
Appendix~\ref{sec:bayes}.

@@

Our broad brush takeaway from Figure~\ref{fig4:repressed_post_full}(C) is
overall of good agreement. There some oddities, in particular the long tails in
the data for Oid, 1~ng/mL, and O2, 0.5~ng/mL. The latter is especially odd since
it extends beyond the tail of the unregulated UV5 distribution. This is a
relatively small number of cells, however, so whether this is a peculiarity of
the experimental data, a statistical fluke of small numbers, or a real
biological effect is unclear. It is conceivable that there is some very slow
timescale switching dynamics that could cause this bimodality, although it is
unclear why it would only appear for specific repressor copy numbers. There is
also a small offset between experiment and simulation for O2 at the higher
repressor copy numbers, especially at 2~ng/mL. We suspect this may be caused by
the very large number of zero mRNA counts for Oid, 2~ng/mL skewing it's partner
datasets through the shared association rate, and so we are not too concerned by
this discrepancy either. None of these fairly minor quibbles cause us to
seriously doubt the overall correctness of our model, which further validates
its use to compare the equilibrium models' binding energies to the
nonequilibrium models' repressor kinetics, as we originally set out to do.

>>>>

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
constitutive promoters, achieves the right balance of complexity and simplicity,
by allowing both Fano factor $\nu>1$, but also by remedying, by design, the
problems of parameter degeneracy that model 4 in
Figure~\ref{fig2:constit_cartoons} suffered~\cite{Razo-Mejia2020}. Does this
stand up to closer scrutiny, namely, comparison to full mRNA distributions
rather than simply their moments? We will test this thoroughly on single-cell
mRNA counts for different unregulated promoters from Jones et.\
al.~\cite{Jones2014}.

@@

For this model the master equation of interest is
Eq.~\ref{eq:poisson_promoter_cme} with repressor set to zero, i.e.,
\begin{equation}
{d\over dt}p_U(m)(t) = 
        rp_U(m-1)(t) 
        - rp_U(m)(t)
        + (m+1)\gamma p_U(m+1)(t) 
        - \gamma p_U(m)(t),
\end{equation}
whose steady-state solution is given by a Poisson distribution with parameter
$\lambda \equiv r / \gamma$~\cite{Sanchez2013}. The goal of our inference 
problem is then to find the probability distribution for the parameter value
$\lambda$ given the experimental data. By Bayes theorem this can be written as
\begin{equation}
p(\lambda \mid D) = {p(D \mid \lambda) p(\lambda) \over p(D)},
\end{equation}
where $D = \{m_1, m_2, \ldots, m_N \}$ are the single-cell mRNA experimental
counts. As is standard we will neglect the denominator $p(D)$ on the right
hand side since it is independent of $\lambda$ and serves only as a
normalization factor.

The steady-state solution for the master equation defines the likelihood term
for a single cell $p(m \mid \lambda)$. What this means is that for a given
choice of parameter $\lambda$, under model 1 of
Figure~\ref{fig2:constit_cartoons}(A), we expect to observe $m$ mRNAs in a
single cell with probability
\begin{equation}
p(m\mid\lambda) = \frac{\lambda^m e^{-\lambda}}{m!}.
\label{eq:poisson_inference010}
\end{equation}
Assuming each cell's mRNA count in our dataset is independent of others, the
likelihood of the full inference problem $p(D\mid\lambda)$ is simply a product
of the single cell likelihoods given by Eq.~\ref{eq:poisson_inference010} above, so
\begin{equation}
p(D\mid\lambda) = \prod_{k=1}^N \frac{\lambda^{m_k}e^{-\lambda}}{m_k!}.
\end{equation}

@@

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is
known as the \textit{conjugate} prior for a Poisson likelihood.
Conjugate priors have the convenient properties that a closed form exists for the posterior distribution $p(\lambda \mid D)$ - unusual in Bayesian inference problems - and the closed form posterior takes the same form as the prior. For
our case of a Poisson distribution likelihood with its
Gamma distribution conjugate prior, the posterior distribution is also a Gamma
distribution~\cite{Gelman2013}. Specifically the two parameters $\alpha'$ and
$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
\frac{1}{N}\sum_{k=1}^N m_k$ for notational convenience, and $N$ is the number
of cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to approximate the
Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
we have inferred our model's one parameter to a precision of order 1\%.

We remind the reader that we began this section claiming that the Poisson model
was ``wrong'' since it could not reproduce features of the data such as a Fano
factor $> 1$. The fact that we obtain such a narrow posterior distribution for our
parameter $P(\lambda \mid D)$ does not equate to the model being adequate to
describe the data. What this means is that given the data $D$, only values
in a narrow range are remotely plausible for the parameter $\lambda$,
but a narrow posterior distribution does not necessarily
mean the model accurately depicts reality.
As we will see later in Figure~\ref{fig:constit_post_full} after
exploring the bursty promoter model, indeed the correspondence
when contrasting the Poisson model with the experimental data is quite poor.

@@

Let us now consider the problem of parameter inference for model five
from~\fig{fig1:means_cartoons}(C). As derived in
Appendix~\ref{sec:gen_fcn_appdx}, the steady-state mRNA distribution in this
model is a negative binomial distribution, given by
\begin{equation}
p(m) = \frac{\Gamma(m+k_i)}{\Gamma(m+1)\Gamma(k_i)}
        \left(\frac{1}{1+b}\right)^{k_i}
        \left(\frac{b}{1+b}\right)^m,
\label{eq:neg_bionom}
\end{equation}
where $b$ is the mean burst size and $k_i$ is the burst rate in units of the
mRNA degradation rate $\gamma$. As sketched earlier, to think of the negative
binomial distribution in terms of an intuitive ``story,'' in the precise
meaning of~\cite{Blitzstein2015}, we imagine the arrival of
bursts as a Poisson process with rate $k_i$, where each burst has a
geometrically-distributed size with mean size $b$.

@@

We carried out Markov-Chain Monte Carlo (MCMC) sampling on the posterior of this
model, starting with the constitutive \textit{lacUV5} dataset
from~\cite{Jones2014}. The resulting MCMC samples are shown in
Figure~\ref{fig:constit_post_full}(A). In contrast to the active/inactive
constitutive model considered in~\cite{Razo-Mejia2020} (nonequilibrium model 4
in Figure~\ref{fig2:constit_cartoons}(A)), this model is well-identified with both
parameters determined to a fractional uncertainty of 5-10\%. The strong
correlation reflects the fact that their product sets the mean of the mRNA
distribution, which is tightly constrained by the data, but there is weak
``sloppiness''~\cite{Transtrum2015} along a set of values with a similar
product.

@@

\fig{fig:constit_post_full}(C) shows the results, plotting the posterior
distribution from individually MCMC sampling all 18 constitutive promoter
datasets from~\cite{Jones2014}. To aid visualization, rather than plotting
samples for each promoter's posterior as in \fig{fig:constit_post_full}(A), for
each posterior we find and plot the curve that surrounds the 95\% highest
probability density region. What this means is that each contour 
encloses approximately 95\% of the samples, and thus 95\% of the probability
mass, of its posterior distribution. Theory-experiment comparisons,
shown in Figure~\ref{figS:ppc_unreg} in Appendix~\ref{sec:bayesian},
display a similar level of agreement between data and predictive samples as for
the bursty model with \textit{lacUV5} in \fig{fig:constit_post_full}(B).

@@

We mention one further comparison, between our inferred parameters and the
work of Chong et.\ al.~\cite{Chong2014}, which is interesting and puzzling.
Beautiful experiments in~\cite{Chong2014} convincingly argue that
supercoiling accumulated from the production of mRNA transcripts is key
in setting the burstiness of mRNA production.
In their model, this supercoiling occurs on the scale of $\sim100$~kb
domains of DNA. This suggests that all genes on a given domain should burst
in synchrony, and that the difference between highly and lowly expressed
genes is the \textit{size} of transcriptional bursts, not the
\textit{time between} bursts.
But here, all burst sizes we infer in Figure~\ref{fig:constit_post_full}(C)
are comparable and burst rates vary wildly. It is not immediately clear
how to square this circle. Furthermore, Figure~7E in~\cite{Chong2014}
reports values of the quantity they label $\beta/\alpha$ and we label
$k^+/k^-$ in model 4 from Figure~\ref{fig2:constit_cartoons}. In contrast
to the findings of~\cite{Razo-Mejia2020}, Chong et.\ al.\ do not find
$k^+/k^-\ll1$ for most of the genes they consider. This begs the question:
is the \textit{galK} chromosomal locus used for the reporter constructs
in~\cite{Razo-Mejia2020} and~\cite{Jones2014} merely an outlier,
or is there a deeper puzzle here waiting to be resolved? Without more
apples-to-apples data we can only speculate, and we leave it as an
intriguing open question for the field.

Despite such puzzles, our goal here is not to unravel the mysterious origins
of burstiness in transcription. Our remaining task in this work is a
determination of the physical reality of equilibrium binding energies in
Figure~\ref{fig1:means_cartoons}, as codified by the equilibrium-nonequilibrium
equivalence of Eq.~\ref{eq:deltaFR_eq_noneq_equiv}.
For our phenomenological needs here model 5 in
Figure~\ref{fig2:constit_cartoons} is more than adequate: the posterior
distributions in Figure~\ref{fig:constit_post_full}(C) are cleanly
identifiable and the predictive checks in Figure~\ref{figS:ppc_unreg}
indicate no discrepancies between the model and the mRNA single-moleucle
count data of~\cite{Jones2014}. Of the models we have considered it
is unique in satisfying both these requirements. So we will happily use
it as a foundation to build upon in the next section when we add regulation.

@@

Before showing the mathematical formulation of our statistical inference model,
we would like to sketch the intuitive structure. The dataset
from~\cite{Jones2014} we consider consists of single-cell mRNA counts data of
nine different conditions, spanning several combinations of three unique
repressor binding sites and four unique repressor copy numbers. We assume that
the values of $k_i$ and $b$ are known, since we have already cleanly inferred
them from constitutive promoter data, and further we assume that these values
are the same across datasets with different repressor binding sites and copy
numbers. In other words, we assume that the regulation of the transcription
factor does not affect the mean burst size nor the burst initiation rate. The
regulation occurs as the promoter is taken away from the transcriptionally
active state when the promoter is bound by repressor. We assume that there is
one unbinding rate parameter for each repressor binding site, and likewise one
binding rate for each unique repressor copy number. This makes our model seven
dimensional, or nine if one counts $k_i$ and $b$ as well. Note that we use only
a subset of the datasets from Jones et.\ al.~\cite{Jones2014}, as discussed more
in Appendix~\ref{sec:bayesian}.

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\},
\end{equation}
where subscripts for dissociation rates $k^-$  indicate the different repressor
binding sites, and subscripts for association rates $k^+$ indicate the
concentration of the small-molecule that controlled the expression of the LacI
repressor (see Appendix~\ref{sec:bayesian}). This is because for this particular
dataset the repressor copy numbers were not measured directly, but it is safe to
assume that a given concentration of the inducer resulted in a specific mean
repressor copy number\cite{Chure2019a}. Also note that the authors
of~\cite{Jones2014} report estimates of LacI copy number per cell rather than
direct measurements. However, these estimates were made assuming the validity of
the equilibrium models in Figure~\ref{fig1:means_cartoons}, and since testing
these models is our present goal, it would be circular logic if we were to make
the same assumption. Therefore we will make no assumptions about the LacI copy
number for a given inducer concentrations.

Having stated the problem, Bayes theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is again the set of all $N$ observed single-cell mRNA counts
across the various conditions. We assume that individual single-cell
measurements are independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding
rates out of $\vect{k}$ for the $j$-th measured cell. The probability
$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
distribution for our bursty model with repression derived in
Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $_2F_1$ is the confluent hypergeometric function of the second kind and
$\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

This likelihood is rather inscrutable. We did not find any of the known
analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
we instead resorted to numerics. One insight we found
was that for very strong or very weak repression, the distribution in
Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
values, except with $k_i$ multiplied by the fold-change
$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the ratio
$k_R^+/k_R^-$ was detectable. But for intermediate repression, the distribution
was visibly broadened with Fano factor greater than $1+b$, the value for the
corresponding constitutive case. This indicates that the repressor rates had
left an imprint on the distribution, and perhaps intuitively, this intermediate
regime occurs for values of $k_R^\pm$ comparable to the burst rate $k_i$. Put
another way, if the repressor rates are much faster or much slower than $k_i$,
then there is a timescale separation and effectively only one timescale remains,
$k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three rates in the problem
are comparable does the mRNA distribution retain detectable information about
them.

Next we specify priors. As for the constitutive model, weakly informative
lognormal priors are a natural choice for all our rates. We found that if the
priors were too weak, our MCMC sampler would often become stuck in regions of
parameter space with very low probability density, unable to move. We struck a
balance in choosing our prior widths between helping the sampler run while
simultaneously verifying that the marginal posteriors for each parameter were
not artificially constrained or distorted by the presence of the prior. The only
exception to this is the highly informative priors we placed on $k_i$ and $b$,
since we have strong knowledge of them from our inference of constitutive
promoters above. All details for our prior distributions are listed in 
Appendix~\ref{sec:bayesian}.

@@

Our broad brush takeaway from Figure~\ref{fig4:repressed_post_full}(C) is
overall of good agreement. There some oddities, in particular the long tails in
the data for Oid, 1~ng/mL, and O2, 0.5~ng/mL. The latter is especially odd since
it extends beyond the tail of the unregulated UV5 distribution. This is a
relatively small number of cells, however, so whether this is a peculiarity of
the experimental data, a statistical fluke of small numbers, or a real
biological effect is unclear. It is conceivable that there is some very slow
timescale switching dynamics that could cause this bimodality, although it is
unclear why it would only appear for specific repressor copy numbers. There is
also a small offset between experiment and simulation for O2 at the higher
repressor copy numbers, especially at 2 and 10~ng/mL.
From the estimate of repressor copy numbers from~\cite{Jones2014}, it is
possible that the repressor copy numbers here are becoming large enough to
partially invalidate our assumption of a separation of timescales between
burst duration and repressor association rate. Another possibility is that
the very large number of zero mRNA counts for Oid, 2~ng/mL is skewing it's partner
datasets through the shared association rate.
None of these fairly minor quibbles cause us to
seriously doubt the overall correctness of our model, which further validates
its use to compare the equilibrium models' binding energies to the
nonequilibrium models' repressor kinetics, as we originally set out to do.
